Space is the Place: Effects of Continuous Spatial Structure on Analysis of Population Genetic Data

### 2019-11-18 ms/review-responses.tex

\pagebreak
\setcounter{page}{1}

%%%%%%%%%%%%%%
\reviewersection{AE}

\begin{quote}
    The manuscript admirably explores a lot of consequences of
    isolation-by-distance in the context of a novel model that is easily amenable
    to forward simulation; however, given that this model may be used in a lot of
    future studies based on the precedent set here, there is some concern about the
    model and its support. Reviewers 2 and 3 highlight this in particular (it
    underlies the main 2 points of reviewer 2's review, and the core of Reviewer
    3's comment), and I agree. Whatever can be done to strengthen the standing of
    this model, and/or connect it to more thoroughly studied models, will be
    helpful for the manuscript. The concern would be that there are peculiarities
    of this model that do not generalize well. A new supplemental section or opener
    to the results section establishing the model more thoroughly would make the
    strongest response.
\end{quote}

This is a good point. There are many ways to go in the quest for demographic realism,
and the fundamental question -- what regulates population density in real populations --
is still very much an open question (or rather, a question with a great many answers).
As highlighted below,
we have included further discussion of our particular choice
of local, density-mediated control on mortality,
along with pointers into the literature, at several places in the manuscript;
see in particular in the Methods when we introduce the model \revreffull{AE}{2},
and \llname{demog_disco}.

@@

\begin{point}{Line 35:}
    Also cite Wilkins and Wakeley, Genetics 2002; Wilkins 2004
\end{point}

\reply{
Done.
\revref
}

\begin{point}{\revref}
    ``Such models have been used extensively in ecological modeling but rarely in population genetics'': Detailing these previous uses via citations and elaboration may help alleviate the major concern about the provenenance of this model and its unique behaviors (see general comments above and R2 and R3 comments).
\end{point}

\reply{
    Good idea --
    we have added some historical discussion and a few more citations to this section,
    \revref. Also see the additional discussion in the Discussion \llname{demog_disco}.
}

@@

\begin{point}{\revref}
    Please describe computation time needed per replicate
\end{point}

\reply{
We have added a figure (Figure \ref{fig:runtimes}) and short discussion \llname{runtimesxn} of run times.
}

\begin{point}{\revref}
    I read the acknowledgement to the Hearth and Falling Sky Brewing with a sense of familiarity in feeling of gratitude to my own favorite cafes and breweries, but I it's not a great precedent for Acknowledgements to be filled this way. Please cut.
\end{point}

@@

\begin{point}{Figure \ref{fig:sumstats}A, \ref{fig:allsfs}:}
    Perhaps more revealing to show on log-log scale?
\end{point}

\reply{
    Good suggestion -- the SFS in Fig \ref{fig:sumstats}A is now on a log-log scale,
    which shows the slight decrease in low frequency SNPs a little better. 
    We've left Fig \ref{fig:allsfs} as-is.
}

\begin{point}{Figure S3:}
    Caption seems to be missing detail 
\end{point}

\reply{
    Thanks for catching this - we have revised this caption to add details including the simulation parameters. (Supplemental Figure \ref{fig:hetmap}).
}

%%%%%%%%%%%%%%
\reviewersection{1}

\begin{quote}
    This study explores biases arising in population-based inference when 1)
    real population samples are coming from spatial habitat with various degree
    of structuring while inference is made assuming random mating population;
    2) imperfect sampling in practice that fails to represent full diversity
    across entire population habitat; 3) phenotypes that vary across geography
    and create spurious associations with genotypes. While earlier studies
    explored the effect of strong structure on population genetic inference and
    GWAS, this work focuses on less extreme scenarios of structuring that
    arises in populations evolving in continuous habitat. By using
    non-Wright-Fisher model, authors simulated chromosome-scale samples from
    populations that evolved in continuous space, and that can model
    environmental factors to create phenotypes varying over space. As a result,
    this study identified spatial structuring scenarios (small neighbourhood
    size $\sim$ 10--100) that coupled with imperfect sampling strategies lead to a
    biased inference of widely used population genetic statistics (altogether
    18 statistics) such as pi (average pairwise sequence differences),
    heterozygosity (and inbreeding coefficient), and IBS tract sharing.
    Accordingly, inference of the effective population size history was also
    strongly affected under these parameter ranges. Finally, the authors use
    their spatial modelling to demonstrate that typical GWAS with PC-based
    correction cannot entirely remove spurious signals of genotype-phenotype
    associations arising from purely environmental factors.  Overall, the
    authors explore an important but often neglected source of bias that can
    affect inference in many population-based studies (in medical genetics,
    evolutionary biology and ecology). This study can be of interest to a
    broader audience of readership, and I have only minor comments to improve
    clarity and increase accessibility for readers:
\end{quote}

We thank the reviewer for their very constructive comments!
Responses follow below:

\begin{point}{}
    When neighbourhood size is small (10-100), the mean number of IBS tracts ${}>2$bp (nIBS as in Table S1) is elevated similar to Wright's inbreeding coefficient, but mean of the distribution of pairwise IBS (mean(IBS)) is decreased. What could be the source of this discrepancy? How exactly mean(IBS) was calculated?
\end{point}

\reply{
    The mean IBS is simply the mean of the vector of lengths of IBS tracts.
    \plr{make sure this is stated in the text somewhere}
    The short answer is that if one splits a chromosome up into more pieces (increasing nIBS),
    the mean length of those pieces must necessarily be smaller.
    At low neighborhood sizes, spatially distant individuals tend to have longer coalescence times,
    leading to more (and hence smaller) IBS tracts.
    This can be seen in Supplemental Figure \ref{fig:allsumstats}.
    Note that skew(IBS) and number of long ($>$1e6) IBS tracts reflect other aspects of the distribution.
}

@@

\begin{quote}
    Battey et al. use spatially explicit population genetic simulations to
    analyze the effects of spatial structure on (i) the estimation of key
    population genetic parameters, in turn used to (ii) make inferences about
    population history, and on (iii) confounding in genome-wide association
    studies (GWAS). I liked the paper a lot. It's interesting, well-written and
    addresses an important question - the effect of spatial population
    structure on population genetic statistics and inference-and I enjoyed
    reading it. The most positive aspects were:

    \begin{enumerate}
        \item It nice to actually see spatially explicit simulations and I'm happy that forward simulation is now fast enough that you can do this sort of thing.
        \item The paper is very clear and well-written, easy to understand the motivation and most of the details. That's not always the case for this sort of paper.
        \item I felt that the section about the effect on GWAS was the most interesting and novel part of the paper and gave me some intuition that I hadn't had before.
    \end{enumerate}

    I don't have any major criticisms. There were a few aspects that I thought might warrant some additional discussion, and a few specific questions below. The general questions I had after reading it were:
\end{quote}

Thanks very much for the encouraging words!

\begin{point}{}
    To what extent are any of the results dependent on the exact method of simulation. There are a number of choices about the exact details of the simulations (e.g. the way the overlapping generations are handled, the edge effects and, particularly, the form of Equation 1 - see below). It's not so much that these are non-standard (since I don't think there is a standard) and they all sort of make sense heuristically, and I was left wondering whether these sorts of choices actually make a difference. Do the authors have some thoughts/intuition/results about that? Given that the results in \autoref{fig:sumstats} seem quite consistent with expectations, I suspect that on some level it doesn't make much difference but then there are intermediate results like \autoref{fig:genparams} which seem a bit counter-intuitive and I wonder if those aspects depend on the simulation scheme.
\end{point}

\reply{
    This is a good point also raised by other reviewers.
    We have added some more discussion of the choice of demographic model
    (\revreffull{AE}{2} and \llname{demog_disco}).
}

@@

\begin{point}{}
    Related to the first point, to what extent are the results qualitatively different to those that would be obtained in a stepping-stone model? My interpretation is that they are actually very similar, but I didn't see whether that was explicitly discussed. In some sense, it's still easier to do large simulations in a stepping-stone model so it would be nice to be reassured that that's still ok.
\end{point}

\reply{
    We have added an appendix comparing our model to a reverse-time stepping stone simulation on several relevant summary statistics \llname{apx:stepping_stone}.
    For some of these statistics (e.g., $\theta_\pi$),
    the behavior of the stepping stone model becomes more similar (although not identical)
    to the continuous model as the resolution of the landscape increases,
    but other statistics seems to be more strongly affected by discretization artifacts.
    We are also curious how all the other analyses we test would be affected using other simulation schemes, but hope to explore that aspect in future work. 
    Furthermore, discrete-deme coalescent simulations are certainly faster,
    but come with significant issues either when neighborhood size is lower than the population of a deme,
    or when demes are small enough that the sample size approaches local $N_e$.
}

@@

\begin{point}{}
    The source of equation \eqref{eqn:p_survival} is not obvious to me. I sort of see how it makes sense, but a little but more intuition or a brief derivation or an illuminating either in the main text or the supplement, would be helpful.
\end{point}

\reply{
    Additional discussion of the model is now provided in the appendix (see "Demographic model") \llname{apx:demographic_model}
}

@@

\begin{point}{}
    The authors use a scaling factor in equation (2) to counteract the increase in fitness of individuals at the edges. Can they provide a figure showing that this is the case. What does ``roughly'' mean on line 164. Perhaps a heatmap of the fitness of individuals across the grid with and without the scaling factor?
\end{point}

\reply{
Good suggestion -- we have added a supplemental figure to show the distribution of individual fitness across the landscape with and without our edge-scaling approach. (Note though that we found the effect subtle and hard to visualize well). 
}

\begin{point}{}
    It would be helpful provide the figure showing that generating mutations during the forward simulations in SLIM is equivalent to applying mutations using msprime on pre-generated trees (line 185)? It sounds like this procedure would underestimate the variance in the number of mutations, since you remove the effect of random generation time. Is this effect small?
\end{point}

\reply{
    We have added a figure showing sample site-frequency-spectra generated from a subset of simulations run with SLiM mutations, and then using msprime to apply mutations to the same tree sequences with our generation-time scaling approach (Figure \ref{fig:slim_v_msp_sfs}). These approaches yield extremely similar spectra so we believe any error in the approximation is quite small. 
}

@@

\end{point}

\reply{
    Similarly here we monitored the population size over time while running SLiM. All simulations maintained asymptotic populations, and none of the runs we started ended because populations crashed.
}

\begin{point}{}
    Third, it is not clear what the relationship between the expected lifespan and the probability of survival is. The expected lifespan, L, is first defined as the inverse of the expected number of offspring produced by a parent. The authors also define the probability of survival of a given individual at a given point in space, $p_i$. Hence, the expected lifespan at a point in space (and time) is the mean of a geometric distribution with parameter $p_i$, i.e., $1/p_i$. Now, it is far from being obvious what the relationship between these two approaches for defining the expected lifespan actually is.
\end{point}

\reply{
   We have added additional discussion of this point to appendix A (see "Demographic Model") \llname{apx:demographic_model}
}

@@

\begin{point}{}
    Also, the web page https://github.com/petrelharp/spaceness does not seem to exist so that I was not able to experiment with the forward-in-time generator used here unfortunately.
\end{point}

\reply{
I apologize -- I had forgotten to set the repo to public. It should work now. 
}

\begin{point}{}
    All in all, more efforts need to be made here in my opinion to show that the forward-in-time simulations generate sensible outcomes. Sensible in terms of the behavior of the population demography at equilibrium (provided such equilibrium indeed exists) along with that of the spatial distribution of individuals. The authors could provide some guarantee of the good behavior of their model as evidenced from simulations using a broad range of parameter values for generating data. Alternatively, they could elect to use the spatial-Lambda-Fleming-Viot model for their simulations, which, in my opinion would seem the most sensible option given that (1) it is possible to run backward-in-time simulations under this model, thereby saving a lot of computation time and (2) it is a well-studied model with good mathematical and biological properties and (3) it is implemented in a publicly available software program (\url{https://github.com/jeromekelleher/discsim}).
\end{point}

\reply{
    This is a good point also raised by other reviewers. We have added an Appendix comparing our model to a stepping-stone \llname{apx:stepping_stone}, rather than the spatial-Labmda-Fleming-Viot, because we think these are the most familiar and widely used class of spatial models. We find that many features of our model are well approximated by stepping-stone models, and that for statistics like $\theta_{\pi}$ the stepping stone model results approach our continuous space model as the number of demes used to describe the landscape increases. 
}

@@

\begin{point}{}
    Definitions and interpretations. There are quite a large number of metrics discussed in Figure 3B, and it's a lot to take in. It might be helpful to have a table with a reminder of what the metric is, its interpretation, and how it is computed.
\end{point}

\reply{
    We have included a table describing the summary statistics in Figure \ref{table:sumstats}.
}

\begin{point}{}
    Notation: ``Offspring disperse a Gaussian-distributed distance away from the parent with mean zero and standard deviation $\sigma$ in both the x and y coordinates. Each offspring is produced with a mate selected randomly from those within distance $3\sigma$, with probability of choosing a neighbor at distance $x$ proportional to $\exp(-x^2/2\sigma^2)$.'' I think x may be overloaded here, or I'm confused. Clarify?
\end{point}

\reply{
	We switched the second instance (referring to the distance among individuals) to $d$. 
}

@@

\reply{
We have rearranged the figure as suggested.

}

\begin{point}{}
    Not sure I follow this example: ``Concretely, an individual at position (x, y) in a 50 × 50 landscape has mean phenotype 100 + 2x/5.''
\end{point}

\reply{
	We have clarified by switching to "the phenotype $p$ for an individual at location (x,y) is then $p=100+2x/5$
}

### 2019-11-19 ms/review-responses.tex

\pagebreak
\setcounter{page}{1}

%%%%%%%%%%%%%%
\reviewersection{AE}

\begin{quote}
    The manuscript admirably explores a lot of consequences of
    isolation-by-distance in the context of a novel model that is easily amenable
    to forward simulation; however, given that this model may be used in a lot of
    future studies based on the precedent set here, there is some concern about the
    model and its support. Reviewers 2 and 3 highlight this in particular (it
    underlies the main 2 points of reviewer 2's review, and the core of Reviewer
    3's comment), and I agree. Whatever can be done to strengthen the standing of
    this model, and/or connect it to more thoroughly studied models, will be
    helpful for the manuscript. The concern would be that there are peculiarities
    of this model that do not generalize well. A new supplemental section or opener
    to the results section establishing the model more thoroughly would make the
    strongest response.
\end{quote}

%\plr{I would generally cut down the quoted bits like the above to only what's essential, but haven't done that yet.}

% \plr{IMPORTANT: don't reorder or delete "points" below - it messes up the automatic numbering!}

### 2019-11-19 ms/spaceness.tex
last(?) pass for reviewer 4

At the beginning of the simulation individuals are distributed uniformly at random on a continuous, square landscape. 
Individuals are hermaphroditic, and each time step, each produces a Poisson number of offspring with mean $1/L$ where $L$ is the expected lifespan.
Offspring disperse a Gaussian-distributed distance away from the parent with mean zero and standard deviation $\sigma$
in both the $x$ and $y$ coordinates.
Each offspring is produced with a mate selected randomly from those within distance $3 \sigma$,
with probability of choosing a neighbor at distance $x$ proportional to $\exp(-x^2 / 2 \sigma^2)$.

@@

In all simulations, the phenotype of each individual
is determined by adding independent Gaussian noise with mean zero \revpoint{1}{5}
to a mean that may depend on spatial position such that the mean of spatially-correlated phenotypes varies by two standard deviations across the landscape.\llabel{phenotypescalingsd}
We then adjust the geographic pattern of mean phenotype to create four spatially autocorrelated environmental influences on phenotype.
In the first simulation of \emph{nonspatial} environments, the mean did not change,
so that all individuals' phenotypes were drawn independently from a Gaussian distribution with mean 110 and standard deviation 10.
Next, to simulate \emph{clinal} environmental influences on phenotype, 
we increased the mean phenotype from 100 on the left edge of the range to 120 on the right edge (two phenotypic standard deviations).
Concretely, an individual at position $(x,y)$ in a $50 \times 50$ landscape has mean phenotype $100 + 2x / 5$.
Third, we simulated a more concentrated ``\emph{corner}'' environmental effect by setting the mean phenotype 
for individuals with both $x$ and $y$ coordinates below 20 to 120 (two standard deviations above the rest of the map). 
Finally, in ``\emph{patchy}'' simulations we selected 10 random points on the map
and set the mean phenotype of all individuals within three map units of each of these points
to 120.

@@

\afterpage{\clearpage}
\begin{figure*}[p]
\centering
\includegraphics[width=\textwidth]{figures/fig_S1_sfs_grid_model_by_sampling.pdf}
\caption{Site frequency spectra for random mating and spatial SLiM models under all sampling schemes.}
\label{fig:allsfs}
\end{figure*}
\FloatBarrier

\llabel{ll:figs3caption}
\begin{figure*}[p]
\centering
\includegraphics[]{figures/het_z_by_ind.pdf}
\caption{Variation in observed heterozygosity (i.e. proportion of heterozygous individuals) in hexagonal bins across the landscape, estimated from a random sample of 200 individuals from the final generation of a simulation with neighborhood size $\approx$25. Values were Z-normalized for plotting.}
\label{fig:hetmap}
\end{figure*}
\FloatBarrier

>>>>

At the beginning of the simulation individuals are distributed uniformly at random on a continuous, square landscape. 
Individuals are hermaphroditic, and each time step, each produces a Poisson number of offspring with mean $1/L$ where $L$ is the expected lifespan.
Offspring disperse a Gaussian-distributed distance away from the parent with mean zero and standard deviation $\sigma$
in both the $x$ and $y$ coordinates.
Each offspring is produced with a mate selected randomly from those within distance $3 \sigma$,
with probability of choosing a neighbor at distance $d$ proportional to $\exp(-d^2 / 2 \sigma^2)$.

@@

In all simulations, the phenotype of each individual
is determined by adding independent Gaussian noise with mean zero \revpoint{1}{5}
to a mean that may depend on spatial position such that the mean of spatially-correlated phenotypes varies by two standard deviations across the landscape.\llabel{phenotypescalingsd}
We then adjust the geographic pattern of mean phenotype to create four spatially autocorrelated environmental influences on phenotype.
In the first simulation of \emph{nonspatial} environments, the mean did not change,
so that all individuals' phenotypes were drawn independently from a Gaussian distribution with mean 110 and standard deviation 10.
Next, to simulate \emph{clinal} environmental influences on phenotype, 
we increased the mean phenotype from 100 on the left edge of the range to 120 on the right edge (two phenotypic standard deviations).
Concretely, the phenotype $p$ for an individual at position $(x,y)$ is $p=100 + 2x / 5$.
Third, we simulated a more concentrated ``\emph{corner}'' environmental effect by setting the mean phenotype 
for individuals with both $x$ and $y$ coordinates below 20 to 120 (two standard deviations above the rest of the map). 
Finally, in ``\emph{patchy}'' simulations we selected 10 random points on the map
and set the mean phenotype of all individuals within three map units of each of these points
to 120.

@@

\begin{figure*}[p]
\centering
\includegraphics[width=\textwidth]{figures/fig_S1_sfs_grid_model_by_sampling.pdf}
\caption{Site frequency spectra for random mating and spatial SLiM models under all sampling schemes.}
\label{fig:allsfs}
\end{figure*}
\FloatBarrier

\begin{figure*}[p]
\centering
\includegraphics[]{figures/het_z_by_ind.pdf}
\caption{Variation in observed heterozygosity (i.e. proportion of heterozygous individuals) in hexagonal bins across the landscape, estimated from a random sample of 200 individuals from the final generation of a simulation with neighborhood size $\approx$25. Values were Z-normalized for plotting.}
\label{fig:hetmap}
\end{figure*}
\FloatBarrier
